% $Author$
% $Date$
% $Revision$

% HISTORY:
% 2007-08-13 - Oscar starts
% 2007-09-01 - Marcus review

%=================================================================
\ifx\wholebook\relax\else
% --------------------------------------------
% Lulu:
	\documentclass[a4paper,10pt,twoside]{book}
	\usepackage[
		papersize={6.13in,9.21in},
		hmargin={.75in,.75in},
		vmargin={.75in,1in},
		ignoreheadfoot
	]{geometry}
	\input{../common.tex}
	\pagestyle{headings}
	\setboolean{lulu}{true}
% --------------------------------------------
% A4:
%	\documentclass[a4paper,11pt,twoside]{book}
%	\input{../common.tex}
%	\usepackage{a4wide}
% --------------------------------------------
    \graphicspath{{figures/} {../figures/}}
	\begin{document}
	\renewcommand{\nnbb}[2]{} % Disable editorial comments
	\sloppy
\fi
%=================================================================
\chapter{Frequently Asked Questions}
\label{app:faq}

\on{These should be *real* (not invented) FAQs.
Here are a few that I have collected.
Check the ST lecture notes for more FAQs.
We should also try to mine more from newbies.}

%=================================================================
\section{Getting started}
\begin{faq}
Where do I get the latest Squeak?
\end{faq}
\answer
\url{ftp.squeak.org/current_development}
\index{download}

\begin{faq}
Which \squeak image should I use with this book?
\end{faq}
\answer
We recommend that you use the image provided on the Squeak by Example web page.
Visit \url{http://SqueakByExample.org} and look for ``Download Squeak'' in the sidebar.
You should also be able to use most other images, but you may find that the hands-on exercises behave differently in surprising ways.

%\begin{faq}
%Where is the Squeak ``Development Image''?
%\end{faq}
%\answer
%\url{www.squeaksource.com/ImageForDevelopers}
%This is a specially-prepared image including lots of useful packages pre-installed for developers.

%=================================================================
\section{Collections}

\begin{faq}
How do I sort an \clsind{OrderedCollection}?
\end{faq}
\answer
Send it the message \mthind{Collection}{asSortedCollection}.

\begin{code}{@TEST}
#(7 2 6 1) asSortedCollection --> a SortedCollection(1 2 6 7)
\end{code}

\begin{faq}
How do I convert a collection of characters to a \clsind{String}?
\end{faq}
\answer
\begin{code}{@TEST}
String streamContents: [:str | str nextPutAll: 'hello' asSet] --> 'hleo'
\end{code}

%=================================================================
\section{Browsing the system}

% See \faqref{omnibrowser}

%\begin{faq}\label{faq:omnibrowser}
%The browser does not look like the one described in the book. What gives?
%\end{faq}
%\answer
%You are probably using an image in which the \ind{OmniBrowser} is installed as the default browser.  You can change the default by clicking on the little menu icon in the top-left corner of the browser window, between the ``\textsf{X}'' and the browser label, and selecting ``Choose new default browser'' from the menu that appears. Select \ct{#Browser} to get the plain old browser.

\begin{faq}
How do I search for a class?
\end{faq}
\answer
\short{b} (browse) on the class name, or \short{f} in the category pane of the browser.
\index{keyboard shortcut!browse it}
\index{keyboard shortcut!find ...}

\begin{faq}
How do I find/browse all sends to super?
\end{faq}
\answer
The second solution is much faster:
\begin{code}{}
SystemNavigation default browseMethodsWithSourceString: 'super'.
SystemNavigation default browseAllSelect: [:method | method sendsToSuper ].
\end{code}
\index{browsing programmatically}
\clsindex{SystemNavigation}

\begin{faq}
How do I browse all super \subind{super}{send}{}s within a hierarchy?
\end{faq}
\answer
\begin{code}{}
browseSuperSends := [:aClass | SystemNavigation default
	browseMessageList: (aClass withAllSubclasses gather: [ :each |
		(each methodDict associations
			select: [ :assoc | assoc value sendsToSuper ])
				collect: [ :assoc | MethodReference class: each selector: assoc key ] ])
	name: 'Supersends of ' , aClass name , ' and its subclasses'].
browseSuperSends value: OrderedCollection.
\end{code}

\begin{faq}
How do I find out which new methods are introduced by a class?  (I.e., not including overridden methods.)
\end{faq}
\answer
Here we ask which methods are introduced by \ct{True}:
\begin{code}{@TEST | newMethods |}
newMethods := [:aClass| aClass methodDict keys select:
	[:aMethod | (aClass superclass canUnderstand: aMethod) not ]].
newMethods value: True --> an IdentitySet(#asBit)
\end{code}

\begin{faq}
How do I tell which methods of a class are abstract?
\end{faq}
\answer
\begin{code}{@TEST | abstractMethods |}
abstractMethods :=
	[:aClass | aClass methodDict keys select:
		[:aMethod | (aClass>>aMethod) isAbstract ]].
abstractMethods value: Collection --> an IdentitySet(#remove:ifAbsent: #add: #do:)
\end{code}

\begin{faq}
How do I generate a view of the \ind{AST} of an expression?
\end{faq}
\answer
Load AST from squeaksource.com. Then evaluate:
\begin{code}{}
(RBParser parseExpression: '3+4') explore
\end{code}
(Alternatively \emph{explore it}.)
\clsindex{RBParser}

\begin{faq}
How do I find all the Traits in the system?
\end{faq}
\answer
\begin{code}{}
Smalltalk allTraits
\end{code}

\begin{faq}
How do I find which classes use traits?
\end{faq}
\answer
\begin{code}{}
Smalltalk allClasses select: [:each | each hasTraitComposition and: [each traitComposition notEmpty ]]
\end{code}
% Should be:
% Smalltalk allClasses select: [:each | each hasTraitComposition ]
% But this is broken for some reason!

%=================================================================
\section{Using Monticello and SqueakSource}

\begin{faq}
How do I load a \ind{Squeaksource} project?
\index{Monticello}
\end{faq}
\answer
\begin{enumerate}
  \item Find the project you want in \url{squeaksource.com}
  \item Copy the registration code snippet
  \item Select \menu{open \go Monticello browser}
  \item Select \menu{+Repository \go HTTP}
  \item Paste and accept the Registration code snippet; enter your password
  \item Select the new repository and \menu{Open} it
  \item Select and load the latest version
\end{enumerate}

\begin{faq}
How do I create a SqueakSource project?
\end{faq}
\answer
\begin{enumerate}
  \item Go to \url{squeaksource.com}
  \item Register yourself as a new member
  \item Register a project (name = category)
  \item Copy the Registration code snippet
  \item \menu{open \go Monticello browser}
  \item \menu{+Package} to add the category
  \item Select the package
  \item \menu{+Repository \go HTTP}
  \item Paste and accept the Registration code snippet; enter your password
  \item \menu{Save} to save the first version
\end{enumerate}

\begin{faq}
How do I extend \ct{Number} with \ct{Number>>>chf} but have Monticello recognize it as being part of my \ct{Money} project?
\end{faq}
\answer
Put it in a method-category named \ct{*Money}.
Monticello gathers all methods that are in other categories named like *package and includes them in your package.

%=================================================================
\section{Tools}

\begin{faq}
How do I programmatically open the \ind{SUnit} \ind{TestRunner}?
\end{faq}
\answer
Evaluate \ct{TestRunner open}.

\begin{faq}
Where can I find the \ind{Refactoring Browser}?
\end{faq}
\answer
Load AST then Refactoring Engine from squeaksource.com:
\url{www.squeaksource.com/AST}
\url{www.squeaksource.com/RefactoringEngine}

\begin{faq}
How do I register the browser that I want to be the default?
\end{faq}
\answer
Click the menu icon in the top left of the Browser window.

%=================================================================
\section{Regular expressions and parsing}

\begin{faq}
How can I work with regular expressions?
\index{regular expression package}
\end{faq}
\answer
\ifluluelse
	{Load Vassili Bykov's RegEx package from: \\
	\url{www.squeaksource.com/Regex.html}}
	{Load Vassili Bykov's RegEx package from:
	\url{www.squeaksource.com/Regex.html}}
\index{Bykov, Vassili}

\begin{faq}
Where is the documentation for the RegEx package?
\end{faq}
\answer
Look at the \menu{DOCUMENTATION} protocol of \ct{RxParser class} in the \menu{VB-Regex} category.

\begin{faq}
Are there tools for writing parsers?
\end{faq}
\answer
Use \ind{SmaCC}\,---\,the Smalltalk Compiler Compiler.
You should install at least SmaCC-lr.13.
Load it from \url{www.squeaksource.com/SmaccDevelopment.html}.
There is a nice tutorial online:
\url{www.refactory.com/Software/SmaCC/Tutorial.html}

\begin{faq}
Which packages should I load from SqueakSource SmaccDevelopment to write parsers?
\end{faq}
\answer
Load the latest version of \ind{SmaCCDev}{}\,---\,the runtime is already there.
(SmaCC-Development is for \squeak 3.8)

%=================================================================
\ifx\wholebook\relax\else\end{document}\fi
%=================================================================

%\begin{faq}
%How do I run a headless image with a file argument?
%\end{faq}
%\answer
%Right now you can't do it with the MacOSX VM.

%\begin{faq}
%How do I find out which methods access the Smalltalk dictionary?
%\end{faq}
%\answer
%???

%\begin{faq}
%How do I get the tree view of an AST?
%\end{faq}
%\answer
%???

%\begin{faq}
%How do I browse all references to a given class?
%\end{faq}
%\answer
%???
