% $Author$
% $Date$
% $Revision$
%=================================================================
% This is the main file for the Squeak By Example book.
% The individual chapters can also be latexed by themselves.
% Note that the \end{document} marker occurs near the
% middle of this file, to leave out additional material
% for a future version of the book.
%=================================================================
\documentclass[a4paper,10pt,twoside]{book}
\usepackage[
	papersize={6in,9in},
	hmargin={.75in,.75in},
	vmargin={.75in,1in},
	ignoreheadfoot
]{geometry}
\input{common.tex}
\setboolean{lulu}{true}
%=================================================================
% Add the path for the figures of each chapter here:
\graphicspath{
	{figures/}
	% {Announcements/figures/}
	{Compiler/figures/}
	% {Concurrency/figures/}
	% {Debugging/figures/}
	{Exceptions/figures/}
	{Installer/figures/}
	% {Magma/figures/}
	{Metaprogramming/figures/}
	{Monticello/figures/}
	{Omnibrowser/figures/}
	{Profiling/figures/}
	{Regex/figures/}
	{Seaside/figures/}
}
%=================================================================
\let\wholebook=\relax
\makeindex
\makeglossary
%=================================================================
% \renewcommand{\nnbb}[2]{} % Disable editorial comments
%=================================================================
\begin{document}
\frontmatter
%=================================================================
\setcounter{page}{1}
\pagestyle{headings}
%=================================================================
%:Inside cover
\author{
	Alexandre Bergel\quad
	Andrew P. Black\quad
	St\'ephane Ducasse\\[1ex]
	Oscar Nierstrasz\quad}
% (to be updated at the end)
\title{\Huge\bf More Squeak by Example\\[1ex]}
\isodate
\date{\emph{Version of \today}}
\maketitle
%=================================================================
%:TOC
\tableofcontents
% \listoffigures
% \listoftables
% \lstlistoflistings
\sloppy % To avoid LaTeX's annoying habit of letting lines stick over the margins!
\mainmatter
%=================================================================
\input{Preface/Preface.tex}
%=================================================================
%:PLANNED for 2d edition
\part{SW Development}
\input{Monticello/Monticello.tex}
\input{Debugging/Debugging.tex}
\input{Profiling/Profiling.tex}
\input{Installer/Installer.tex}

\part{Frameworks}
\input{Seaside/Seaside.tex}
\input{Magma/Magma.tex}
\input{Regex/Regex.tex}
\input{Omnibrowser/Omnibrowser.tex}

\part{Language}
\input{Exceptions/Exceptions.tex}
% \input{Concurrency/Concurrency.tex} % ON: seems to be broken for now
\input{Announcements/Announcements.tex}
\input{Metaprogramming/Metaprogramming.tex}

%=================================================================

% \printglossary
\bibliographystyle{jurabib}
\bibliography{scg}

\printindex

\end{document}

%=================================================================
%:Planned for later releases

% The following chapters are planned for later releases.



%:Compiler
\input{Compiler/Compiler.tex}


%=================================================================

\chapter{Inheritance}

\on{or: The darker side of inheritance ?}

Include use and misuse of Inheritance.

\on{should be about idioms, not design!}

Prefer abstract superclasses.

Include the stuff that never gets talked about, like weak subclasses and variable subclasses.

Why it's impossible to subclass SmallInteger.

\ab{programming idioms involving inheritance.  singleton, initialization, set... methods for instance initialization, self class and self species rather than using explicit class names.  More as they appear.  }

\chapter{Exceptions}

Stef sez: see the test cases

%	For exceptions
%	--------------
%	On Oct 26, 2006, at 9:23, stephane ducasse wrote:

%	BTW oscar if you want to have fun look at the test for the exception in Squeak.
%	They are excellent. I was planning to use them to illustrate exception
%	(and understand myself deeply) in a separate chapter.

%	I mean

%		on:do:
%		
%	Look at the classes in the category Exceptions-Tests.
%	You have all the possible scenario on what we can do with exception in smalltalk
%	and this is impressive.

%	ExceptionTests  uses ExceptionTester to check all the possible combinations and
%	semantical points. I just read everything once just to learn.
%	Now I would like to remember it and I guess that writing it down would be the best
%	(No time for that so far)

%	Stef
\chapter{Concurrency}
Processes and friends.

\chapter{Dependency Mechanisms}
Stef sez: we need on overview on publish/subscribe and the dependency mechanisms in squeak.

\chapter{Common Errors}
The last chapter of Smalltalk with Style is a good source.

\chapter{Leftover stuff}
Time.

\section{XML processing :)}

\section{How make an image self executable withSqueak}
We save an image with our main application window already opened, and unwanted things disabled (halos, alt-., menus, etc). Additionally, we have a class-side \#startUp: method to initialize what's necessary on startup, like re-reading config files etc.

The image is started automatically when double-clicking  "plopp.exe", or rather a short-cut in the program menu placed there by the installer. This exe is nothing more than a renamed Squeak.exe, with a different icon patched in using a resource editor. There is a VM config file "plopp.ini", which most importantly contains the path to the image to load (see http://minnow.cc.gatech.edu/squeak/3274).

The exe along with the needed plugins, the image and other data files is installed in the "programs" directory by a regular Windows installer (InstallShield).

It's really not too different from other applications, which nowadays also consist of an executable and several additional files. That the actual application logic in Squeak's case is not contained in the executable but in a "data file" is of no concern to the end user.

Note that this is an FAQ item (http://minnow.cc.gatech.edu/squeak/3563). Feel free to put my explanation on the Swiki, I haven't checked carefully if a similar answer is already there.

- Bert -
%=================================================================

% \printglossary
\bibliographystyle{jurabib}
\bibliography{scg}

\printindex

\end{document}

%:===> END OF ADDITIONAL MATERIAL <===
% \end{document} % NB: The actual end is further up
%=================================================================
%%% Local Variables:
%%% coding: utf-8
%%% mode: latex
%%% TeX-master: t
%%% TeX-PDF-mode: t
%%% ispell-local-dictionary: "english"
%%% End:
