% $Author: ducasse $
% $Date: 2009-08-24 10:17:33 +0200 (Mon, 24 Aug 2009) $
% $Revision: 28563 $

%% 2011-09-11 - Migrated to PharoBox: svn checkout https://XXX@scm.gforge.inria.fr/svn/pharobooks/PharoByExampleTwo-Eng

%=================================================================
\ifx\wholebook\relax\else
% --------------------------------------------
% Lulu:
	\documentclass[a4paper,10pt,twoside]{book}
	\usepackage[
		papersize={6.13in,9.21in},
		hmargin={.75in,.75in},
		vmargin={.75in,1in},
		ignoreheadfoot
	]{geometry}
	\input{../common.tex}
	\setboolean{lulu}{true}
% --------------------------------------------
% A4:
%	\documentclass[a4paper,11pt,twoside]{book}
%	\input{../common.tex}
%	\usepackage{a4wide}
% --------------------------------------------
    \graphicspath{{figures/} {../figures/}}
	\begin{document}
\fi
%=================================================================
%\renewmessage{\nnbb}[2]{} % Disable editorial comments
\sloppy

%=================================================================
%\renewcommand{\nnbb}[2]{#2} % Disable editorial comments

\chapter{Coral: Scripting in Pharo}

Have you ever dreamed to write your shell scripts in Smalltalk? 
Coral is then for your. 


\section{Getting Coral}

You can either use a prebuild image that you can find on the Pharo build server @@ here @@ or build
the Coral environment.

\begin{itemize}

\item Take a Pharo 1.3 image (here, for example:
\url{https://pharo-ic.lille.inria.fr/hudson/view/Pharo/job/Pharo Core 1.3/} \sd{probably wrong now.}

\item Open the image and load the package ÒConfigurationOfCoralÓ from ÒCoralÓ SqueakSource repository. The fastest way is to evaluate:


\begin{code}{}
Gofer new 
	squeaksource: 'Coral'; 
	package: 'ConfigurationOfCoral'; load.
\end{code}

\item Then, evaluate (this line is written on the comment of the class \ct{ConfigurationOfCoral})

\begin{code}{}
(ConfigurationOfCoral project version: '0.4') load
\end{code}

\end{itemize}
At this point, Coral is installed from the Pharo perspective. But better install the helper files as explained in the next section.


\section{Helper Files}

Now to help you launching it. Coral can generate for you some scripts:

\begin{code}{}
Coralnstaller generateCoralScript.
Coralnstaller generateCoralDebugScript
\end{code}

This generates two shells that you can use as follows:

\begin{code}{}
./script.sh CoralScript.cst [arguments to Coral script]
\end{code}

\begin{itemize}
\item \ct{coral.sh} runs pharo in headless mode and quits when the Coral script is ended.
\item \ct{coralDebug.sh} runs pharo with UI and does not quit even if the Coral script is ended
\end{itemize}


Do not forget to give them the execute right:
\begin{code}{\sd{we should do that from coral!}}
chmod +x coral.sh coralDebug.sh
\end{code}

Note that these scripts remember the locations and names of your virtual machine and of your Coral image. YouÕll have to generate them again if any of these information changes.

 
The last thing you can do is to generate all examples given with Coral:

\begin{code}{}
CoralInstaller generateAllExamples
\end{code}

 This message  takes all selectors in CoralScriptExamples class in the Òscript examplesÓ category, and generate a file for each selector, which content will be the return value (a string containing a Coral script) of the evaluation of the selector.
 
 
At that point, you can execute any example. For example:
 
\begin{code}{}
./coral.sh scriptHappyFace.cst
\end{code}
 
\section{How to write Coral scripts?}
 
Coral syntax is almost the same as the Pharo syntax. 
 There are just two things added:
\begin{itemize}
\item  \ct{[ ]}: anything you put into square brackets will be executed.

\item Method bodies are surrounded by \ct{[ ]}.

For example here is the way to define a method on point which takes one argument
\begin{code}{}
Point>>myAreaScaled: scale
	[
	^ self x * self y * scale
	]
\end{code}
The only things that changes are \ct{Point>>} and \ct{[} after the parameter list and the closing one \ct{]}. 


For adding a method on class side, use: \ct{aClass class>>aSelectorName}
  \end{itemize}
  
  
  
\paragraph{Examples}
  
\begin{code}{\sd{we should use real examples}}
[ 
	Transcript show: 'This line is in two (unnamed) square brackets, so it is executed'. 
]

Object>>uselessMethod 
	[ 
	"This method does nothing but is added to the instance side of Object" 
	]
	
Object class>>anotherUselessMethod: aString 
	[
	"This method does nothing too, but it is added on class side. And it takes a paremeter". 
	]
\end{code}
  

\subsection{Class Definition} 
  
Since the definition of a subclass is quite long, there is a small syntax added

\begin{code}{}
Object < newSubClass
\end{code}

A hint to remember the order: if you put a pipe it looks like the UML triangle of generalization: \ct{<|}.


Take a look to the CoralScriptWriting class. It provides methods to access to the arguments given to the script, by such a command:

\begin{code}{}
./coral.sh myScript.cst foo -option 42
\end{code}


\section{Dealing with Files}

\section{Deamons}

\section{}


%=========================================================
\ifx\wholebook\relax\else
   \bibliographystyle{jurabib}
   \nobibliography{scg}
   \end{document}
\fi
%=========================================================




%%% Local Variables: 
%%% coding: utf-8
%%% mode: latex
%%% TeX-master: Lint.tex
%%% TeX-PDF-mode: t
%%% End:

